\documentclass[11pt]{article}
\usepackage{macros}
\singlespacing
\begin{document}

\section{Externality}

Kevin is a graduate student in economics and is endowed with a dollar amount
$y_A$ per day.  Francisco is not an economist and is endowed with dollar amount
$y_B$ per day. It costs Kevin \$4 an hour to produce economics research. Let $x$
denote the number of hours per day that Kevin devotes to research. Kevin's
research gives Francisco trading signals, and thus Francisco also has a
preference over Kevin's research. Kevin's and Francisco's utilities are $u_A,
u_B$, respectively:

\begin{align*}
  u_A(x, y_A) &= \begin{cases}
  y_A + 8x - x^2/2 & x \le 8 \\
  y_A + 32 & o.w.
\end{cases} \\
  u_B(x, y_B) &= \begin{cases}
  y_B + 12x - x^2/2 & x \le 12 \\
  y_B + 72 & o.w.
\end{cases}
\end{align*}
    
\begin{enumerate}

% Problem 1
\item What are the marginal rates of substitution between Francisco and Kevin?

\begin{sol}
In order to find the marginal rate of substitution of Kevin, we can differentiate
his ``revenue" function with respect to $x$ treating $y_A$ as a function of $x$
given by the amount of $y_A$ necessary to keep utility constant. Then

$$\arraycolsep=1.4pt\def\arraystretch{1.5} 0 = \left\{ \begin{array}{ll} \frac{dy_A}{dx} + 8 - x & x \le 8 \\ \frac{dy_A}{dx} & x \ge 8 \end{array} \right. \implies -\frac{dy_A}{dx} = \left \{ \begin{array}{ll} 8-x & x \le 8 \\ 0 & x \ge 8 \end{array} \right..$$

Similarly for Francisco,

$$\arraycolsep=1.4pt\def\arraystretch{1.5} 0 = \left\{ \begin{array}{ll} \frac{dy_B}{dx} + 12 - x & x \le 12 \\ \frac{dy_B}{dx} & x \ge 12 \end{array} \right. \implies -\dfrac{dy_B}{dx} = \left\{ \begin{array}{ll} 12 - x & x \le 12 \\ 0 & x \ge 12 \end{array} \right..$$

\end{sol}

% Problem 2
\item How much will Kevin produce if he doesn't know that Francisco exists.

\begin{sol}
If Kevin doesn't even know that Francisco exists, he will produce music to the
point where the marginal benefit of it equals its cost. In other words, he
solves
$$\dfrac{d}{dx} U_A(x,y_A) = 4 \implies 8 - x = 4 \implies \boxed{x=4}.$$
Producing any quantity of $x > 8$ is not ideal because he derives no marginal
benefit to music at that point, but still accrues the cost.
\end{sol}

% Problem 3
\item Suppose that Kevin produces the amount of research in 2 and Francisco has
found a way to consume Kevin's research by downloading it for free from ArXiv.
Kevin has no way to prevent Francisco's free-riding. How much consumer surplus
do Francisco and Kevin obtain?

\begin{sol}
Kevin produces 4 units of research. For the $x$th unit of research, Francisco
derives a benefit of $12-x$. Thus, Francisco's total consumer surplus is given
by,

$$\int_0^4 12- x\, dx = 12x - \dfrac{x^2}{2} \bigg|_0^4 = \boxed{40}.$$

On the other hand, Kevin's benefit from the $x$th unit of research is given by
$(8-x) - 4$ which is the benefit of the research minus its cost. Integrating
over all four units gives,
\[ \int_0^4 4 - x \, dx = 4x - \frac{x^2}{2} \bigg|^4_0 = 16 - 8 = \boxed{8}.\]

\end{sol}


% Problem 4
\item Pareto efficient allocations are characterized by the relation, ``sum of
marginal rates of substitution equals marginal cost of production.'' What is the
Pareto efficient volume of research for Kevin to produce? What is the total
surplus at this level of research?

\begin{sol}
We first calculate the sum of the marginal rates of substitution

\[ -\dfrac{dy_A}{dx} - \dfrac{dy_B}{dx} = 
\left\{ \begin{array}{ll} 
  20 - 2x & \text{if } x \le 8 \\
  12 - x & \text{if } 8 \le x \le 12 \\
  0 & \text{if } 12 \le x
\end{array} \right. \]

and equating this value to 4 gives us that $\boxed{x=8}$ is the Pareto efficient
volume of research for Kevin to produce. The total consumer surplus for the
$x$th unit of music is the sum of the marginal rates of substitutions minus the
cost of production, and since 8 units of music are being created, the total
consumer surplus is,
\[ \int_0^8 (20 - 2x) - 4 \, dx = 16x - x^2 \bigg|^8_0 = 128 - 64 = \boxed{64}.\]
\end{sol}

% Problem 5
\item Now suppose that Kevin has found a way to protect his research from
Francisco's free-loading. From now on, Francisco will have to pay $p$ dollars
for every unit of research he downloads. If Kevin chooses the price $p$ that
maximizes his ``profit from Francisco'' (revenue from Francisco, minus cost of
production), what price will he charge and how much research will he produce?
What will be Kevin and Francisco's consumer surpluses?

\begin{sol}
First, we must analyze how much music Francisco will buy at price $p$, which is
the $x$ such that

$$p = 12 - x \implies x = 12 - p.$$

Then, this means that Kevin will work to maximize

$$y_A + 8x - \dfrac{x^2}{2} - 4x + x(12 - x) \implies x = \dfrac{16}{3}$$

and consequently he will set the price equal to $\frac{20}{3}$. Thus Francisco's
consumer surplus is

$$\int_0^{16/3} 8 - x\, dx = \dfrac{256}{9}$$

and similarly, Kevin's consumer surplus is

$$\int_0^{16/3} \dfrac{32}{3} - x \, dx = \dfrac{128}{3}$$

as well.

\end{sol}


\end{enumerate}
<<<<<<< HEAD
\section{Spicy Cobb--Douglas}

Suppose there are two goods $x,y$ and utility \[
u(x,y) = \begin{cases}
     (x-\bar a)^\alpha (y-\bar b)^{1- \alpha} & x > \bar a, y > \bar b \\
     -\infty & o.w.
\end{cases}
\]
with $\bar a, \bar b > 0.$ Let $w$ denote wealth and assume $p_x \bar a + p_y \bar b < w$.  
\begin{enumerate}
    \item Derive the Marshallian demand for $x,y$. 
    \begin{sol}
        Observe that one must purchase $\bar a$ units of $x$ and $\bar b$ units of $y$. Assume that one has already done so. Let $\tilde w = w - (p_x \bar a + p_y \bar b)$. Let $\tilde x = (x-\bar a)$ and $\tilde y = (y-\bar b)$. Then the problem is \[
        \max_{\tilde x, \tilde y} \, \tilde x ^{\alpha} \tilde y^{1-\alpha}\text{ subjected to } p_x \tilde x + p_y\tilde y = \tilde w,
        \]
        which is a vanilla Cobb--Douglas problem.
        
        Therefore \[
        \tilde x = x-\bar a = \frac{\alpha \tilde w}{p_x}\quad \tilde y = y-\bar b = \frac{(1-\alpha)\tilde w}{p_y},
        \]
        yielding the Marshallian demand \[
        x = \bar a + \frac{\alpha \tilde w}{p_x} \quad y = \bar b + \frac{(1-\alpha) \tilde w}{p_y}.
        \]        
    \end{sol}
    
    \item Derive the Hicksian demand functions for $x$ and $y$ from the cost minimization problem. 
    \begin{sol}
        The Cobb--Douglas first order conditions are \[
        \frac{\tilde y}{\tilde x} = \frac{1- \alpha}{\alpha}\frac{p_x}{p_y} \implies \tilde y = A \tilde x,
        \]
        for some $A$.
        
        Then \[
        A^{1-\alpha} \tilde x = \bar U \implies \tilde x = \pr{\frac{\alpha p_y}{(1- \alpha) p_x}}^{1-\alpha} \bar U.
        \]
        and \[
        \tilde y = A x = A^{\alpha} \bar U = \pr{\frac{(1- \alpha)p_x}{\alpha p_y}}^\alpha \bar U.
        \]    
    \end{sol}
    \item Show that the own-price Slutsky equation for good $x$ holds for this utility function.
    \begin{sol}
        We wish to show that \[
        \frac{dx}{dp_x} = \diff{h_x}{p_x} - x \diff{x}{w}.
        \]
        Note that \[\frac{dx}{dp_x} = \alpha \frac{\bar a p_x - w + \bar a p_x + \bar b p_y}{p_x^2} = -\alpha \frac{w - \bar b p_y}{p_x^2} \]
        \[
        \diff{h_x}{p_x} = - (1-\alpha) \pr{\frac{\alpha p_y}{(1- \alpha) p_x}}^{1-\alpha} \frac{\bar U}{p_x},
        \]
        where \[\bar U = \pr{\frac{\alpha \tilde w}{p_x}}^\alpha \pr{\frac{(1-\alpha) \tilde w}{p_y}}^{1-\alpha}.\]
        Then \[
        \diff{h_x}{p_x}
= -(1-\alpha) \alpha \tilde w \frac{1}{p_x^2}.
        \]
        Lastly \[
        x\diff{x}{w} = \pr{\bar a + \frac{\alpha \tilde w}{p_x}}\frac{\alpha}{p_x}.
        \]
        Then \begin{align*}
        \text{RHS} &= -(1-\alpha) \alpha \tilde w \frac{1}{p_x^2} - \pr{\bar a + \frac{\alpha \tilde w}{p_x}}\frac{\alpha}{p_x} \\
        &= -\alpha \tilde w p_x^{-2} + \alpha^2 \tilde w p_x^{-2} - \bar a \alpha p_x p_x^{-2} - \alpha^2 \tilde w p_x^{-2} \\
        &= -\alpha p_x^{-2} \pr{\tilde w + \bar a p_x} = -\alpha p_x^2 (w - \bar b p_y) = \text{LHS}.
        \end{align*}
        \end{sol}
        \item Using your Slutsky decomposition from the previous part, can you say which demand function, Marshallian or Hicksian, is more sensitive to price changes? Is this true for all goods and all utility functions (briefly)?
        \begin{sol}
            Note that the income effect $-x\diff{x}{w}$ is negative. Thus Marshallian demand is more sensitive. This is not true of all goods and utility functions (consider inferior goods). 
        \end{sol}

\end{enumerate}
=======

\section{Cobb--Douglas}
>>>>>>> 34a609692d9b5bcc3e06d638a760ee3742f09bbd

\section{Vertical Monopolies}
There is a unit mass of identical price-taking consumers. Each consumer has income $y$ and consumes widgets $x$ and some num\'eraire good $x_0$, with utility function \[
u(x, x_0) = x_0 + \frac{1}{\sigma+1} x^{\sigma+1}, \quad -1 < \sigma < 0
\]
The price of $x$ is $p$ and the price of $x_0$ is $1$.
\begin{enumerate}
    \item Assume that $y$ is large so that $x_0, x$ have positive demand. Derive the Marshallian demand for $x$ as a function of $p$. Given $p$, derive the consumer surplus in good $x$.
    \begin{sol}
        The first order conditions are \[
        x^{\sigma} = \lambda p \quad \lambda = 1.
        \]
        Thus $x = p^{1/\sigma}$ and $x_0 = y - p^{\frac{\sigma + 1}{\sigma}}$, assuming $y > p^{\frac{\sigma + 1}{\sigma}}$. 
        
        The consumer surplus is \[
        \text{CS} = \int_{p}^{\infty} t^{1/\sigma}\,dt = \frac{1}{1+1/\sigma}t^{1/\sigma + 1}\Bigg \rvert_{p}^\infty = \frac{-\sigma}{\sigma + 1} p^{1 + 1/\sigma}.
        \]
    \end{sol}
    
     \begin{sol}\textbf{[Alternative Solution]}
          Recognize that the Marshallian demand for a quasilinear utility function $u(a,b) = a + v(b)$ with $p_a=1, p_b=p$ is \[b = (v')^{-1}(p).\]
              \end{sol} 
    \item What's the elasticity of demand?
    \begin{sol}
        \[
        \diff{\log x}{ \log p} = \frac{1}{\sigma}. 
        \]
    \end{sol}
    \item 
    Assume the consumers purchase positive amounts of $x$.
    The consumers purchase $x$ from a monopolist retailer, who sets the price $p$. Suppose the retailer faces constant marginal cost $c_R$, solve for the profit-maximizing $p$ as a function of $c_R$. Derive the retailer's profit $\pi_R$. 
    \begin{sol}
         The monopolist maximizes \[
         \max_p \, p^{1+1/\sigma} - c_R p^{1/\sigma} 
         \]
         Thus \[
         \frac{1+\sigma}{\sigma} p^{1/\sigma} - \frac{1}{\sigma}c_R p^{\frac{1-\sigma}{\sigma}} = 0 \implies p = \frac{1}{1+\sigma} c_R. 
         \]
         
         The retailer's profit is \[
         \pi_R = \pr{\frac{1}{1+\sigma} c_R }^{1 + 1/\sigma} - c_R\pr{\frac{1}{1+\sigma} c_R }^{1/\sigma}
         \]
         
     \end{sol}
     \begin{sol}\textbf{[Alternative Solution]}
          Recognize that the monopolist sets price with the markup rule \[
          \text{Price} = \pr{\frac{1}{1+1/\epsilon}} \cdot \pr{\text{Marginal Cost}} = \frac{1}{1+\sigma} c_R,
          \]
          where $\epsilon < 0$ is the elasticity of demand.
      \end{sol} 
     \item Assume the monopolist purchases $x$ from a monopolistic wholesaler, who sets $c_R$ and faces constant, exogenous marginal cost $c_W$. Solve for the profit-maximizing $c_W$. Derive the wholesaler's profit $\pi_W$ and the sum of the wholesaler and retailer's profits $\pi_W + \pi_R$.
     \begin{sol}
     Note that \[
     x = \pr{\frac{1}{1+\sigma} c_R }^{1/\sigma}
     \]
         The wholesaler solves \[
         \max_p \, c_R \pr{\frac{1}{1+\sigma}c_R }^{1/\sigma} - c_W \pr{\frac{1}{1+\sigma} c_R}^{1/\sigma}
         \]
    By the problem above, we have that \[c_R = \frac{1}{1+\sigma} c_W.\] 
    
    Let $A = \frac{1}{1+\sigma}$. We have \begin{align*}
        \pi_W &= A^{1/\sigma} \pr{c_R^{1+1/\sigma} - c_W c_R^{1/\sigma}}
        \\
        &=  A^{1/\sigma} \pr{A^{1+1/\sigma} c_W^{1+1/\sigma} - A^{1/\sigma}c_W c_W^{1/\sigma}} \\
        &= A^{2/\sigma} \pr{A-1}c_W^{1+1/\sigma}\\
        \pi_W + \pi_R &= \pr{A^2 c_W}^{1 + 1/\sigma} - A c_W \pr{A^2 c_W}^{1/\sigma} + A^{2/\sigma} \pr{A-1}c_W^{1+1/\sigma} \\
        &= A^{2+2/\sigma} c_W^{1+1/\sigma} - A^{1+2/\sigma} c_W^{1+1/\sigma} + A^{1 + 2/\sigma} c_W^{1+1/\sigma} - A^{2/\sigma}c_W^{1+1/\sigma} \\
        &= A^{2/\sigma} (A+1)(A-1) c_W^{1+1/\sigma}.
    \end{align*}
    
     \end{sol}
     
     \item What is the total surplus $\text{CS} + \pi_W + \pi_R$?
     \begin{sol}
     \begin{align*}
     \text{Total Surplus} &= A^{2/\sigma} (A+1)(A-1) c_W^{1+1/\sigma} - \frac{\sigma}{\sigma + 1} \pr{A^2 c_W}^{1 + 1/\sigma} \\
     &= c_W^{1+1/\sigma} \pr{A^{2/\sigma} (A+1)(A-1) - \sigma A^{3+2/\sigma}}
     \end{align*}
     \end{sol}
     
     \item Suppose the retailer and wholesaler merge into a monopoly that charges $p_{\text{merged}}$ to the consumers and takes $c_W$ marginal cost. Find the optimal $p_{\text{merged}}$ and total surplus.  
     \begin{sol}
         $p_{\text{merged}} = Ac_W.$ The monopoly's profit is \[
         \pr{Ac_W}^{1+1/\sigma} - c_W (Ac_W)^{1/\sigma} = A^{1/\sigma}(A-1) c_W^{1+1/\sigma}. 
         \]
         \begin{align*}
         \text{Total Surplus} &=    A^{1/\sigma}(A-1) c_W^{1+1/\sigma} - \frac{\sigma}{\sigma+1} \pr{Ac_W}^{1+1/\sigma} \\
         &= c_W^{1+1/\sigma} \pr{A^{1/\sigma}(A-1) -\sigma A^{2+1/\sigma}}
         \end{align*}
         
     \end{sol}
     \item Suppose $\sigma = -1/2$. Should the total-surplus-maximizing Federal Trade Commission let the retailer and wholesaler merge?
     
    \begin{sol}
             Consider \begin{align*}
             \frac{\text{TS}_\text{separate}}{\text{TS}_{\text{merged}}} &= \frac{ c_W^{1+1/\sigma} \pr{A^{2/\sigma} (A+1)(A-1) - \sigma A^{3+2/\sigma}}}{c_W^{1+1/\sigma} \pr{A^{1/\sigma}(A-1) -\sigma A^{2+1/\sigma}}}\\
             &= A^{1/\sigma} \pr{A + \frac{A-1}{-\sigma A^2 + A -1}} \\
             \end{align*}
            Plugging in $\sigma = -1/2$, $A = 2$, we have that the ratio equals $\frac{1}{4}\pr{2 + \frac{1}{2 + 2 - 1}} = \frac{7}{12} < 1$. Therefore the FTC should let the firms merge. 
         \end{sol}     
    
    \item (Bonus) Prove that the total surplus under the merged firm is larger for all $\sigma \in (-1,0)$. 
    [Hint: Use a version of Bernoulli's inequality: $(1+t)^r \ge rt$ for $t > -1$, $r \le 0$] 
    \begin{sol}
    Observe that \[
    -\sigma A^2 + A - 1 = \frac{-\sigma + (1+\sigma) - (1+\sigma)^2}{(1+\sigma)^2} = \frac{- 2 \sigma - \sigma^2 }{(1+\sigma)^2} = \frac{-\sigma(\sigma+2)}{(1+\sigma)^2}
    \]
        Consider \begin{align*}
             \frac{\text{TS}_\text{separate}}{\text{TS}_{\text{merged}}} &= \frac{ c_W^{1+1/\sigma} \pr{A^{2/\sigma} (A+1)(A-1) - \sigma A^{3+2/\sigma}}}{c_W^{1+1/\sigma} \pr{A^{1/\sigma}(A-1) -\sigma A^{2+1/\sigma}}}\\
             &= A^{1/\sigma} \pr{A + \frac{A-1}{-\sigma A^2 + A -1}} \\
              \\
             &= A^{1/\sigma} \pr{A + \frac{(1+\sigma)^2}{-\sigma (\sigma + 2)} \frac{-\sigma}{1+\sigma}} \\
             &= A^{1+1/\sigma} \pr{1 + \frac{(1+\sigma)^2}{\sigma+2}} \\
             &=  A^{1+1/\sigma} \pr{\frac{\sigma^2 + 3 \sigma + 3}{\sigma + 2}}.
             \end{align*}
             By the inequality $(1+x)^r \ge 1 + rx$, we have 
             \[
             A^{1+1/\sigma} = \frac{1}{(1+\sigma)^{1+1/\sigma}} \le \frac{1}{2+\sigma}.
             \]
             Thus \[
             \frac{\text{TS}_\text{separate}}{\text{TS}_{\text{merged}}} \le \frac{\sigma^2 + 3 \sigma + 3}{(\sigma+2)^2} = \frac{\sigma^2 + 3 \sigma + 3}{\sigma^2 + 3 \sigma + (4 + \sigma)} < 1
             \]
             for any $\sigma \in (-1,0)$. 
    \end{sol}
    \item True/False/Uncertain: Mergers always hurt the consumer.
\end{enumerate}
\end{document}
